---
title: "Percent_Mining_Calculator"
author: "Matthew Ross"
date: "11/12/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(sf) #Amazingly simple tidy GIS R package
library(mapview) #Interactive mapping of sf objects
library(mapedit) #Interactive editing of spatial data
library(rayshader) #Lovely 3D rendering platform
library(tidyverse) #Good 'ol tidyverse (dplyr, readr, more)
library(elevatr) #R access to mapzen (dynamic downloading of DEMs)
library(raster) # Name says it all, rasters in R
library(whitebox)
library(stars) # A bridge to the future of raster (spatiotemporal objects)
library(rgl) 
library(rfigshare)
library(USAboundaries)
library(tmap)
knitr::opts_chunk$set(echo = TRUE)

# Run once!
#devtools::install_github("giswqs/whiteboxR")
```


In 2018 Andrew Pericack and others published a paper mapping the extent of 
mountaintop mining activities in West Virginia, Virginia, Kentucky, and Tennessee.
In this [paper](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0197758), we only do this analysis from an areal point of view. So if
you wanted to know how much mining was upstream of any given point in the region
we couldn't tell you. But using the amazing `whitebox` tools, we can use R
to calculate the % mining of any river in the region and then extract that to
any arbitrary point. 

First we need to download the cumulative mining dataset that Andrew made which 
is [here](https://figshare.com/s/1383008730d98258f96a) and an elevation dataset
which we will use to calculate watershed metrics.

# Data acquisition and organization

## Cumulative mining from Pericack et al., 2018
```{r}

#Create a data repo if one doesn't exist.
if(all(grepl('data',list.files()))){
  dir.create('data')
} else{print('data folder already exists')}

  

#Download datasets with a check so it doesn't get downloaded twice


cume_file <- 'data/cume.tif'

if(file.exists(cume_file)){
  print('file already downloaded')
}else{
  download.file('https://ndownloader.figshare.com/files/11429579?private_link=1383008730d98258f96a',destfile=cume_file,method='libcurl',mode='wb')
  }


```


## Check out this raster

It's huge so we will subset it to one county in WV and shrink it to 
90m resolution from 30m

```{r}

cume <- raster('data/cume.tif')


boone <- us_counties(resolution='high',states='WV') %>%
  filter(name == 'Boone') %>%
  st_transform(st_crs(cume))
  

cume_boone <- crop(cume,boone) %>%
  aggregate(.,3)


mapview(cume_boone)
```


## Download relevant elevation data

Our watershed analyses will be based on elevation data which we will 
get from elevatr

```{r}
elev <- get_elev_raster(cume_boone,z=10) %>%
  crop(.,cume_boone) %>%
  projectRaster(.,cume_boone)

#Save this elev data for whitebox
writeRaster(elev,'data/elev.tif',overwrite=T)

#Create a raster with same extent as elev but where all values are set
#to 1

full <- setValues(elev,1) 
empty <- setValues(elev,0)
writeRaster(full,'data/full.tif',overwrite=T)
writeRaster(empty,'data/empty.tif',overwrite=T)
writeRaster(cume_boone,'data/cume_boone.tif',overwrite=T)

```

## Analyze DEM using whitebox

Whitebox is an amazing tool with strong documentation [here](https://jblindsay.github.io/wbt_book/available_tools/hydrological_analysis.html)

```{r}


raw_file <- 'data/elev.tif'

#Fill single cell pits (for hydrologic correctness)
wbt_breach_single_cell_pits(raw_file,'data/breach1.tif')
#Breach depressions (better option that pit filling according to whitebox docu
#mentation) The flat_increment bit was tricky for me and needed to be tuned.
wbt_breach_depressions('data/breach1.tif',
                             'data/breached.tif',
                             flat_increment=.1)

#D8 pointer (what is a pointer? a flow direction grid)
wbt_d8_pointer('data/breached.tif','data/d8_pntr.tif')

#D8 flow accumulation (raster cells fully drain in 1 of 8 directions)
wbt_d8_flow_accumulation('data/breached.tif',
                     'data/d8_flow.tif',
                     out_type='catchment area',
                     log=F)


wbt_d8_mass_flux(dem='data/breached.tif',
                 loading='data/cume_boone.tif',
                 efficiency='data/full.tif',
                 absorption='data/empty.tif',
                 output='data/mine_cume.tif')



# Dinf flow accumulation (Flow can be partially partitioned to many cells )
wbt_d_inf_flow_accumulation('data/breached.tif',
                        'data/d_inf_flow.tif',
                        out_type='catchment area',
                        log=T)



#Extract the stream network with a threshold of 11
#This threshold should be adjusted
#wbt_extract_streams('data/d8_flow.tif',output='data/streams.tif',threshold=11)

```


## Watershed plots

First let's just look at the flow accumulation output

```{r}
d8 <- raster('data/d8_flow.tif')

tm_shape(d8) + 
  tm_raster(style='log10')

```


```{r}
mcume <- raster('data/mine_cume.tif')*(90*90)

tm_shape(mcume) + 
  tm_raster()


d_streams <- d8

d_streams[d_streams < 10000] <- NA

pmine <- mcume/d_streams*100
pmine[pmine < 1] <- NA

tm_shape(pmine)+
  tm_raster(style='log10')

```

